---
title: "Homework1.Rmd"
output:
  pdf_document: default
  html_document: default
date: "2024-10-10"
---

```{r setup, include=FALSE, tidy.opts=list(width.cutoff=60)}
### Packages and Data Setup

# Load the necessary packages
library(tidyverse)
library(palmerpenguins)
library(gridExtra)

# Set global chunk options for displaying multiple plots on the same page
knitr::opts_chunk$set(fig.show = "hold", fig.width = 15, fig.height = 9)

```

## Task 1: Improved Plots

```{r echo=FALSE}
# Load the mpg dataset as a Data Frame
df <- as.data.frame(mpg)

# Check the Data Frame
head(df)
```

```{r echo=FALSE}
# Create a frequency table for the 'drv' variable
freq_table <- as.data.frame(table(mpg$drv))

# Calculate relative frequency and percentage
freq_table$rel_Freq <- round(freq_table$Freq / sum(freq_table$Freq), 2)
freq_table$Percentage <- round(freq_table$rel_Freq * 100, 2)

# Rename the columns for clarity
colnames(freq_table) <- c("drv", "Freq", "rel_Freq", "Percentage")

# Print the frequency table
print(freq_table)
```

```{r echo=FALSE}
# Create a bar chart
ggplot(freq_table, aes(x = drv, y = Freq, fill = drv)) +
  geom_bar(stat = "identity") +
  labs(title = "Frequency of Drive Types", x = "Drive Type (drv)", y = "Frequency") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

```{r echo=FALSE}
# Create a pie chart
ggplot(freq_table, aes(x = "", y = Freq, fill = drv)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("4" = "red", "f" = "green", "r" = "blue")) +
  labs(title = "Distribution of Drive Types", fill = "Drive Type (drv)") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

```{r echo=FALSE}
# Simple histogram with custom breaks
hist(mpg$hwy, 
     breaks = seq(10, 45, by = 5),main = "Histogram of Highway Mileage",
     xlab = "Highway",
     ylab = "Frequency",
     col = "yellow")

```

```{r echo=FALSE}
# Histogram with x-axis labeled from 12 to 45 in steps of 2

hist(mpg$hwy, 
     
     breaks = seq(12, 45, by = 2),main = "Histogram of Highway Mileage",
     xlab = "Highway",
     ylab = "Frequency",
     col = "orange")

```

```{r echo=FALSE}
# Simple Boxplot

boxplot(mpg$hwy, horizontal=TRUE,main="Boxplot of Highway Mileage",xlab="Highway",
        col="lightblue")

```

```{r echo=FALSE}
# Scatter plot with adjusted y-axis limits, no grid, and white points with black borders
ggplot(df, aes(x = displ, y = hwy)) +
  geom_point(color = "magenta", fill = "lightblue", shape = 21) +
  labs(title = "Relationship between Engine Displacement and Highway Mileage", 
       x = "Displacement", 
       y = "Highway Mileage (mpg)") +
  scale_y_continuous(breaks = seq(15, 45, by = 5), limits = c(0, NA), expand = c(0, 0)) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

```{r echo=FALSE}
# boxplot showing the association between highway mileage and drive system 
ggplot(df, aes(y = drv, x = hwy)) +
  geom_boxplot(fill = "lightblue", color = "black", outlier.shape = 21, 
               outlier.fill = "white", outlier.color = "black") +
  labs(title = "Association between Drive System and Highway Mileage", 
       x = "hwy", y = "drv") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")  # Center the title
  ) +
  scale_x_continuous(breaks = seq(15, 45, by = 5), limits = c(10, 45))

```

```{r echo=FALSE}
# Stacked bar chart
barplot(table(mpg$drv),col = c("green","blue","cyan"),
        main="Stacked bar chart",
        xlab="drv",
        ylab="hwy")
 
```

```{r echo=FALSE}
# Stacked bar chart showing the distribution of vehicle classes within each drive type
ggplot(df, aes(x = drv, fill = class)) +
  geom_bar(position = "stack", color = "black") +
  labs(title = "Distribution of Vehicle Classes by Drive System", 
       x = "Drive System (drv)", 
       y = "Count", 
       fill = "Vehicle Class") +
  scale_fill_manual(values = c("2seater" = "red", 
                             "compact" = "yellow", 
                             "midsize" = "limegreen", 
                             "minivan" = "cyan", 
                             "pickup" = "blue", 
                             "subcompact" = "purple", 
                             "suv" = "magenta"))+
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

```

```{r echo=FALSE}
# Clustered bar chart showing association between drv and class
df_complete <- as.data.frame(table(mpg$drv, mpg$class))
colnames(df_complete) <- c("drv", "class", "n")

ggplot(df_complete, aes(x = drv, y = n, fill = class)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  labs(title = "Association between categorical variables", 
       x = "Drive System (drv)", y = "Count", fill = "Vehicle Class") +
  scale_fill_manual(values = c("2seater" = "red", 
                               "compact" = "yellow", 
                               "midsize" = "limegreen",
                               "minivan" = "cyan", 
                               "pickup" = "blue", 
                               "subcompact" = "purple", 
                               "suv" = "magenta")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

```{r echo=FALSE}
# Stacked bar chart with proportions
df_propor <- df_complete
df_propor$proportion <- df_propor$n / ave(df_propor$n, df_propor$drv, FUN = sum)

ggplot(df_propor, aes(x = drv, y = proportion, fill = class)) +
  geom_bar(stat = "identity", position = "fill", color = "black") +
  labs(title = "Proportion of Vehicle Classes by Drive System",
       x = "Drive System (drv)", 
       y = "Proportion", 
       fill = "Vehicle Class") +
  scale_y_continuous(labels = scales::percent) +
scale_fill_manual(values = c("suv" = "magenta", 
                             "subcompact" = "purple", 
                             "pickup" = "blue", 
                             "minivan" = "cyan", 
                             "midsize" = "limegreen", 
                             "compact" = "yellow", 
                             "2seater" = "red"))+

  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

```{r echo=FALSE}
# Mosaic Plot
mosaicplot(table(mpg$drv,mpg$class),
           col = c("red","yellow","green","lightgreen","lightblue","blue","magenta"),
           main="Mosaic plot",las = 0.5,xlab="",ylab="")
 
```

## Task 2: Association between engine displacement and highway mileage

```{r echo=FALSE}
# Scatter plot to show the association between engine displacement and highway mileage -> EMPR_03_Visualization1_AS2024.pdf / pages 4,7 and 14
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(shape = fl, color = class)) +
  geom_smooth(method = "lm", se = F) +
  labs(title = "Association between Engine Displacement and Highway Mileage",
       x = "Engine Displacement (liters)",
       y = "Highway Mileage (mpg)",
       color = "Car Class",
       shape = "Fuel Type") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

## Task 3: Comparison of geom_point() and geom_count()

```{r echo=FALSE}
# Scatter plot using geom_point -> EMPR_03_Visualization1_AS2024.pdf / pages 2,4 and 5
point_plot <- ggplot(data = mpg, aes(x = cty, y = hwy)) + 
  geom_point(alpha = 0.6) +  # Add slight transparency for overlapping points
  labs(title = "Scatterplot using geom_point", x = "City MPG", y = "Highway MPG") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# Count plot using geom_count
count_plot <- ggplot(data = mpg, aes(x = cty, y = hwy)) + 
  geom_count(color = "blue", alpha = 0.6) +  # Color to enhance density visibility
  labs(title = "Count Plot using geom_count", x = "City MPG", y = "Highway MPG") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# Display plots side by side
grid.arrange(point_plot, count_plot, ncol = 2)

```

### Explanation

For Task 3, we use geom_point() and geom_count() to explore the relationship between cty (city MPG) and hwy (highway MPG) in the mpg dataset. In the scatter plot with geom_point() (as suggested in the slides on visualizing distributions),each point represents an observation, which can obscure dense areas if points overlap. Adding slight transparency (alpha) mitigates this and reveals overlapping data points. In contrast, geom_count() changes the size of each point based on its count,providing a clear indication of where data points are densest, a recommendation seen in the slides for visual clarity when values overlap. Overall, geom_point() is ideal for datasets with low overlap, while geom_count() effectively highlights density in datasets with repeated values. These enhancements meet the general assignment criteria for readability and clarity in data presentation.

## Task 4: Penguins

```{r echo=FALSE}
# Load the penguins dataset
data(penguins)

# Original (copy of the homework)
ggplot(data = penguins, aes(fill = island, x = species)) +
geom_bar(aes(y = after_stat(prop))) +
  labs(title = "Orginal Plot") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# Improved version with absolute counts displayed over proportionally filled bars -> EMPR_04_Visualization2_AS2024.pdf / page 8
p <- ggplot(data = penguins, aes(x = species, fill = island)) + 
  geom_bar(position = "stack") +
  geom_text(stat = "count",
            aes(label = after_stat(count)),
            position = position_stack(vjust = 0.5),
            color = "white",
            vjust = 1.5) +
  labs(title = "Count of Penguin Species by Island with Proportion",
       x = "Penguin Species",
       y = "Count") +
  scale_fill_manual(values = c("Biscoe" = "steelblue",
                               "Dream" = "coral", 
                               "Torgersen" = "lightgreen")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# Display the plot
p
```

### Problems with the original plot

In the original plot, you can only see which penguin species are present on each island. For example, on Torgersen Island, there are only Adelie penguins, whereas Biscoe Island has both Gentoo and Adelie penguins. However, it does not give a sense of where there are larger populations or the total number of birds on each island. That’s why we added the total counts to each bar in our updated plot. Additionally, the proportions are now displayed accurately, which enhances the clarity of the visualization. For instance, you can immediately see that while Adelie penguins are the most common species, Biscoe Island has the largest overall penguin population.

## Task 5: Diamond

```{r echo=FALSE, fig.width=16, fig.height=10, warning=FALSE, message=FALSE}
# Load the diamonds dataset
data(diamonds)

# Basic bar chart - Clarity distribution grouped by diamond cut
bar_plot <- ggplot(diamonds, aes(x = cut, fill = clarity)) + 
  geom_bar(position = "dodge") +
  labs(title = "Diamond Cut by Clarity (Grouped Bar Chart)", 
       x = "Diamond Cut", y = "Count") +
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank(),
        plot.margin = unit(c(1, 1, 1, 1), "cm")) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# Stacked bar chart - Clarity distribution stacked within each cut category
stacked_bar_plot <- ggplot(diamonds, aes(x = cut, fill = clarity)) + 
  geom_bar(position = "stack") +
  labs(title = "Diamond Cut by Clarity (Stacked Bar Chart)", 
       x = "Diamond Cut", y = "Count") +
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank(),
        plot.margin = unit(c(1, 1, 1, 1), "cm")) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# Pie chart - Proportions of clarity within each cut category
pie_chart <- ggplot(diamonds, aes(x = "", fill = clarity)) + 
  geom_bar(width = 1, position = "fill") +
  coord_polar("y") +
  facet_wrap(~cut) +
  labs(title = "Proportions of Clarity by Diamond Cut (Pie Chart)") +
  theme_minimal() +
  theme(axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.text = element_blank(), 
        legend.position = "bottom", 
        legend.title = element_blank(),
        plot.margin = unit(c(1, 1, 1, 1), "cm")) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# Another pie chart using polar coordinates
pie_chart1 <- ggplot(diamonds) + 
  geom_bar(aes(cut)) + 
  coord_polar() + 
  labs(title = "Doughnut Chart", 
       x = "Cut", y = "Count") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# Display the charts in a 2x2 grid with adjusted spacing
grid.arrange(bar_plot, stacked_bar_plot, pie_chart, pie_chart1)
```

### Comparison

For this task, we created four visualizations to explore the distribution of clarity within each cut category in the diamonds dataset. The grouped bar chart displays clarity levels side-by-side for each cut category, allowing straightforward comparisons of diamond counts by clarity, as seen on slide 10. The stacked bar chart also shows the same data but stacks clarity levels within each cut category. This visualization highlights the relative contributions of each clarity level to the total counts, enhancing our understanding of the dataset's composition. Next, the faceted pie chart provides insights into clarity proportions for each cut, displaying one pie chart per cut. This format illustrates the distribution of clarity levels clearly, utilizing polar coordinates as advised on slides 12 and 13. Finally, the alternative pie chart view transforms a basic bar chart into a pie chart using polar coordinates. While less detailed than the faceted version, it offers a quick overview of clarity proportions across all cut categories. In summary, the grouped bar chart excels for clear comparisons, while the stacked bar chart and both pie charts offer valuable insights into the relative proportions of clarity, each showcasing its unique strengths.

### **Workload**

We sat down at the beginning and divided up the tasks. The first thing we did was set up a Git repository so that we could easily collaborate and continuously track each other’s progress. We organized the tasks so that Fabian took on tasks 1 and 2, while Samuel handled tasks 3-5. This clear division helped us focus and avoid overlapping efforts. After completing our initial assignments, we held a short meeting to discuss the status and share updates on our progress. This check-in proved beneficial, as it highlighted some areas that needed further refinement. Although a few items were still incomplete, we felt that we were moving in the right direction and understood what remained to be done. To ensure the quality of each other’s work, we decided that Fabian would review and make corrections to Samuel's tasks, and vice versa. This mutual review process not only helped catch errors but also facilitated a better understanding of each other's approach. By the end of the session, we felt more confident about our progress and looked forward to wrapping up the remaining tasks.
